Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-09-05 Manuscript/manuscript.tex
Update results and discussion, binding affinity section is not ready yet

Before proceeding to ion binding affinity studies, we quantify
the response of headgroup order parameters to the amount of 
bound charge by using mixtures of monovalent cationic surfactants
and POPC~\cite{scherer89}. The amount of bound charge per PC 
in these systems is 
given by the molar fraction of cationic surfactants, because essentially all surfactants 
locate in the lipid bilayers. Comparison of headgroup order parameter
changes between simulations and experiments 
as a function of mole fraction of cationic surfactants 
can be then used to validate the sensitivity of
lipid headgroup to bound charge in simulation models.

@@

The headgroup order parameter changes as a function of mole fraction
of cationic surfactant dihexadecyldimethylammonium bromide (C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$) 
is compared between experiments~\cite{scherer89} and simulations in
Fig. \ref{OrderParameterCHANGESsurf}. 
Approximately linear decrease of the order parameters is observed
in simulations and experiments with increasing amount of cationic
surfactants up to mole fractions of $\sim$50\%,
as expected from Eq. \ref{OPchangeEQ}.
The slope is, however, too steep in Lipid14 model indicating that 
the headgroup order parameters are too sensitive to a bound charge.
The ECC-lipids model gives a slope in very good agreement with experiments
for the $\alpha$ segment, while the slope is slightly
underestimated for the $\beta$ segment.
%This has to be taken into account when interpreting the
%head group order parameter response to CaCl$_2$ concentration in Lipid14 model.
%observed in  Ref. \citenum{catte16}, see also Fig. \ref{OrderParameterCHANGESnewMODELS}.
%The present comparison of head group order parameter changes 
%suggests that the observed overestimated response of all models to 
%increasing CaCl$_2$ concentration in \cite{catte16} 
%can be explained at least in part by a high sensitivity of the head group order parameter response
%to the bound charge.
\todo{SAMULI: We could calculate the slopes from simulations, but I am not sure if
we would actually learn anything useful from this.}

Also the effect of bound cations on the
headgroup P-N vector angle with respect to the membrane normal
is shown in Fig.~\ref{OrderParameterCHANGESsurf}.
The headgroup orients more parallel to the membrane
normal with an increasing amount of bound cations, as expected~\cite{seelig87}. 
The orientation effect is more pronounced in Lipid14 model than 
in ECC-lipids model, which is in line with the order parameter 
results and reduced charge--dipole interactions in the ECC-lipid model.
Since the response of $\alpha$-order parameter is in very good agreement
with experiments in ECC-lipid model, we can suggest a relation between
order parameter and P-N vector angle changes. The data gives
$\Delta \Theta_{\rm P-N}=?? \Delta S_{\rm CH}$, which can be used
to estimate P-N vector angle changes from experimental results
for $\alpha$ order parameters. 
\todo{SAMULI: I think that we should establish the relation between 
order parameter and P-N vector angle changes by using ECC-lipids model.
This would be useful for people measuring aplha order parameter
changes from experiments.}

@@

The linear decrease of the headgroup order parameters, quantified in previous 
section, can be used to measure the amount of bound aqueous cations in lipid 
bilayers~\cite{seelig87,catte16}. The headgroup order parameter response 
to NaCl and CaCl$_2$ concentrations from experiments (DPPC \cite{akutsu81} 
and POPC \cite{altenbach84}) and different simulation models for POPC
are shown in Fig. \ref{OrderParameterCHANGESnewMODELS}.
%The simulation results of Lipid14 with \AA{}qvist ions~\cite{aqvist90} (from Ref. \citenum{catte16})
%are shown together with the results from Lipid14 simulation
%with ions by Dang~\cite{smith94,chang1999,dang2006} and from ECC-lipids simulations with ECC-ions.

Experimentally NaCl has a negligible effect on the headgroup order parameters
with submolar concentrations due to the very weak binding of Na$^+$
in PC bilayers \cite{akutsu81,catte16}. 
While almost all the available simulation models overestimate cation
binding in PC lipids, the weak binding of Sodium ions was reproduced by
Lipid14 model when simulated with \AA{}qvist ions~\cite{catte16}.
However, the headgroup order parameter response to CaCl$_2$ concentration was
overestimated with the same combination of force field parameters~\cite{catte16}.
Changing to ion models by Dang et al.~\cite{smith94,chang1999,dang2006} or
ECC-ions~\cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014}
with more realistic bulk behaviour were used did not improve the results
for the CaCl$_2$ interactions with Lipid14 model, as seen in 
in Figs. \ref{OrderParameterCHANGESnewMODELS} and \ref{??} (in SI), respectively.
\todo{Add OP-response of Lipid14+ECC-ions plot in SI}.
Together with extensive data in previous work \cite{catte16}, 
the results suggest that CaCl$_2$ binding behaviour in PC bilayers
cannot be corrected by only improving the ion models.

>>>>

Before proceeding to the ion binding affinity studies, we quantify
the response of headgroup order parameters to the amount of 
bound charge by using mixtures of monovalent cationic surfactants
and POPC~\cite{scherer89}. The amount of bound charge per PC 
in these systems is given by the molar fraction of cationic 
surfactants, because essentially all surfactants locate in the 
lipid bilayers. Experimental data for these systems can be used to validate 
the sensitivity of lipid headgroup order parameters
to the amount of bound charge in simulations.

@@

The headgroup order parameter changes with increasing amount of 
cationic surfactant dihexadecyldimethylammonium bromide (C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$) 
is compared between experiments~\cite{scherer89} and simulations in
Fig. \ref{OrderParameterCHANGESsurf}. 
Approximately linear decrease is observed
in simulations and experiments with increasing amount of cationic
surfactants up to mole fractions of $\sim$50\%,
as expected from Eq. \ref{OPchangeEQ}.
The slope is, however, too steep in Lipid14 model indicating that 
the headgroup order parameters are too sensitive to a bound charge.
The ECC-lipids model gives a slope in very good agreement with experiments
for the $\alpha$ segment, while the slope is slightly
underestimated for the $\beta$ segment.
%This has to be taken into account when interpreting the
%head group order parameter response to CaCl$_2$ concentration in Lipid14 model.
%observed in  Ref. \citenum{catte16}, see also Fig. \ref{OrderParameterCHANGESnewMODELS}.
%The present comparison of head group order parameter changes 
%suggests that the observed overestimated response of all models to 
%increasing CaCl$_2$ concentration in \cite{catte16} 
%can be explained at least in part by a high sensitivity of the head group order parameter response
%to the bound charge.
\todo{SAMULI: We could calculate the slopes from simulations, but I am not sure if
we would actually learn anything useful from this.}

The headgroup P-N vector angle with respect to the membrane normal
is also shown in Fig.~\ref{OrderParameterCHANGESsurf}
as a function of cationic surfactant mole fraction.
The headgroup orients more parallel to the membrane
normal with increasing amount of bound cations, as expected~\cite{seelig87}. 
The effect is more pronounced in Lipid14 than 
in ECC-lipids model, which is in line with the order parameter 
results and the reduced charge--dipole interactions in the ECC-lipid model.
The response of $\alpha$-order parameter to bound positive charge
in ECC-lipid model is in good agreement with experiments.
Therefore we use the data in Fig.~\ref{OrderParameterCHANGESsurf}
to suggest a linear relation between $\alpha$-carbon order parameter
and average P-N vector angle changes $\Delta \Theta_{\rm P-N}=?? \Delta S_{\rm CH}^{\alpha}$.
This can be used to estimate the change in P-N vector angle
from experiments measuring changes in $\alpha$-carbon order parameter. 
\todo{SAMULI: I think that we should establish the relation between 
order parameter and P-N vector angle changes by using ECC-lipids model.
This would be useful for people measuring aplha order parameter
changes from experiments.}
For example, $\Delta S_{\rm CH}^{\alpha}=0.018$ measured due to the
addition of 66.8 $\mu$M concentration of Mellittin in bulk water \cite{kuchinka89}
gives $\Delta \Theta_{\rm P-N} = 2.5^{\circ}$.
\todo{I looked this roughly from the figure, should be calculated from 
the actual relation if we decide to put this.}

@@

The binding affinity of aqueous cations in lipid 
bilayers can be measured by using the headgroup 
order parameters, because they decrease linearly 
with the bound positive charge~\cite{seelig87,catte16}. 
The headgroup order parameter responses 
to aqueous NaCl and CaCl$_2$ concentrations 
are shown in Fig. \ref{OrderParameterCHANGESnewMODELS}
from experiments (DPPC \cite{akutsu81} 
and POPC \cite{altenbach84}) and different simulation models for POPC

%The simulation results of Lipid14 with \AA{}qvist ions~\cite{aqvist90} (from Ref. \citenum{catte16})
%are shown together with the results from Lipid14 simulation
%with ions by Dang~\cite{smith94,chang1999,dang2006} and from ECC-lipids simulations with ECC-ions.

Negligible changes of the headgroup order parameters are measured
with submolar concentrations of NaCl due to the very low affinity of Na$^+$
in PC bilayers \cite{akutsu81}. While Na$^+$ binding and the related 
headgroup order parameter changes were overestimated in
almost all the available simulation models, the low affinity
and negligible order parameter changes were reproduced by Lipid14 
model when simulated with \AA{}qvist ions~\cite{catte16}.
However, the same combination of force field parameters
overestimated the headgroup order parameter response to CaCl$_2$ 
concentration, which was the case also in all other models tested in
Ref.~\citenum{catte16}.
Using ion model by Dang et al.~\cite{smith94,chang1999,dang2006} or
ECC-ions~\cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014}
with more realistic bulk behaviour did not improve the results
for the CaCl$_2$ interactions with Lipid14 model, as seen in 
in Figs. \ref{OrderParameterCHANGESnewMODELS} and \ref{??} (in SI), respectively.
\todo{Add OP-response of Lipid14+ECC-ions plot in SI}.
The results support the conclusion of the previous work \cite{catte16}
that improvements also in lipid models are needed to 
correctly describe cation binding in PC bilayers.
